---
title: "traceAsthma: Methods & Reproducible Pipeline"
subtitle: "Sex-Stratified Epigenetic Regulatory Network Analysis of Asthma"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 8
    fig-height: 5
    geometry: margin=1in
---

## Overview

This report documents the **traceAsthma** pipeline: a two-stage causal framework tracing
**SNP → PRS → methylation → eQTM → transcription factor → TRACE score**, applied under a
sex-stratified / disease-stratified analytical design. It combines the written methods with
the R code that produces each pipeline stage, so the report and the analysis stay in sync —
edit the code, re-render, and the figures/tables update automatically.

```{r setup}
library(tidyverse)
library(MatrixEQTL)
library(bnlearn)
library(igraph)
# library(traceAsthma)  # uncomment once installed: devtools::install_github("Eskezeia/traceAsthma")

set.seed(2025)  # fixed seed for reproducibility, per Methods

install.packages("tinytex")
tinytex::install_tinytex()

tinytex::tinytex_root()

```

## Pipeline overview

```{mermaid}
flowchart TD
    A["SNP genotypes"] --> B["Polygenic Risk Score (PRS)"]
    B --> C["DNA Methylation"]
    C --> D["eQTM Discovery<br/>(5 regression models, cis/trans)"]
    D --> E["Tier 1: Pooled BH-FDR &lt; 0.05"]
    E --> F["Tier 2: Effect-size heterogeneity<br/>(Wald test, p_diff &lt; 0.05)"]
    F --> G["Transcription Factor mapping"]
    G --> H["Bayesian Network<br/>(hill-climbing, bootstrap τ = 0.85)"]
    H --> I["Key-Driver Scoring<br/>(out-degree, betweenness, eigenvector)"]
    I --> J["TRACE Score"]
    J --> K["External GWAS/eQTL Validation<br/>(MAGMA, TWAS/S-PrediXcan)"]
    K --> L["Evidence Tiering<br/>A: Convergent | B: Epigenetic-only | C: Genetic-only"]
```

## Stage 1 — eQTM discovery

Five parallel regression models are fit per dataset (discovery, replication) to characterize
disease-shared, disease-specific, disease-adjusted, and disease-discordant CpG–gene
associations. Cis-eQTM pairs are within 1 Mb; all others are trans.

$$Y = \beta_0 + \beta_1 X + \sum_{k=1}^{K} \alpha_k Z_k + \varepsilon \quad \text{(Model 1: disease-combined)}$$

$$Y = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 (X \times D) + \sum_{k=1}^{K} \alpha_k Z_k + \varepsilon \quad \text{(Model 4: disease-interaction)}$$

```{r eqtm-discovery}
#| label: fig-eqtm-volcano
#| fig-cap: "Effect size vs. significance across eQTM models (placeholder — replace with real MatrixEQTL output)"

# --- Placeholder: replace with your actual MatrixEQTL call ---
# eqtm_result <- Matrix_eQTL_engine(
#   snps = methylation_slice,
#   gene = expression_slice,
#   cvrt = covariates_slice,
#   output_file_name = NULL,
#   pvOutputThreshold = 0.05,
#   useModel = modelLINEAR,
#   verbose = TRUE
# )

# Simulated placeholder data for scaffold purposes
set.seed(1)
sim_eqtm <- tibble(
  beta = rnorm(2000, 0, 0.4),
  neg_log10_p = -log10(runif(2000, 1e-8, 1)),
  model = sample(c("Combined", "Asthma-specific", "Control-specific",
                    "Interaction", "Adjusted"), 2000, replace = TRUE)
)

ggplot(sim_eqtm, aes(beta, neg_log10_p, color = model)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  labs(x = expression(hat(beta)), y = expression(-log[10](p)),
       title = "eQTM effect sizes across the five regression models") +
  theme_minimal()
```

## Tier 1 / Tier 2 filtering

Precision filtering removes pairs with $SE_i > Q_{0.95}\{SE_{(AS)}, SE_{(HC)}\}$. Remaining
Model 2/3/4 *p*-values are pooled and BH-corrected to define **Tier 1** ($q < 0.05$); a
two-sample Wald test on the asthma- vs. control-specific effect estimates then defines
**Tier 2** ($p_{diff} < 0.05$):

$$z_{diff} = \frac{\hat\beta_{AS} - \hat\beta_{HC}}{\sqrt{SE_{AS}^2 + SE_{HC}^2}}$$

```{r tiering}
#| label: fig-tier-counts
#| fig-cap: "Number of CpG–gene pairs retained at each filtering tier (placeholder)"

tier_counts <- tibble(
  tier = factor(c("All tested", "Precision-filtered", "Tier 1 (BH-FDR)", "Tier 2 (Wald)"),
                levels = c("All tested", "Precision-filtered", "Tier 1 (BH-FDR)", "Tier 2 (Wald)")),
  n = c(48210, 41032, 3187, 962)  # placeholder counts — wire up to real pipeline output
)

ggplot(tier_counts, aes(tier, n)) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = scales::comma(n)), vjust = -0.4) +
  labs(x = NULL, y = "CpG–gene pairs") +
  theme_minimal()
```

## Stage 2 — Bayesian network & key-driver scoring

Tier 2 cis pairs define a directed whitelist $W$ for hill-climbing structure search
(`bnlearn`), stabilized via bootstrap edge confidence ($R = 200$, $\tau = 0.85$). Key drivers
are ranked by a composite centrality score:

$$\text{Score}(v) = Z[d^{out}(v)] + Z[b(v)] + Z[e(v)]$$

```{r bayesian-network}
#| label: fig-bn-graph
#| fig-cap: "Consensus Bayesian network structure (placeholder graph)"

# --- Placeholder: replace with your bnlearn hill-climbing + boot.strength() output ---
g <- sample_gnp(30, 0.08, directed = TRUE)
V(g)$score <- scale(degree(g, mode = "out") + betweenness(g) + eigen_centrality(g)$vector)[,1]

plot(g,
     vertex.size = 6 + 4 * scales::rescale(V(g)$score, to = c(0, 1)),
     vertex.color = "#41ab5d", vertex.label = NA,
     edge.arrow.size = 0.3, main = "Consensus network (placeholder)")
```

```{r key-drivers}
#| label: tbl-key-drivers

tibble(
  gene = paste0("Gene_", 1:10),
  out_degree = round(rnorm(10, 5, 1.5), 2),
  betweenness = round(rnorm(10, 20, 6), 2),
  eigenvector = round(runif(10, 0.4, 1), 2)
) |>
  mutate(score = round(scale(out_degree)[,1] + scale(betweenness)[,1] + scale(eigenvector)[,1], 2)) |>
  arrange(desc(score)) |>
  knitr::kable(caption = "Top candidate key drivers (placeholder — replace with real Score(v) ranking)")
```

## External GWAS/eQTL validation

Candidate key drivers are cross-referenced against a GWAS-linked gene set built from GABRIEL/TAGC
summary statistics via MAGMA and TWAS/S-PrediXcan (GTEx weights), then tested for enrichment:

$$Z_{TWAS} = \frac{w^T Z_{GWAS}}{\sqrt{w^TRw}}$$

```{r evidence-tiering}
#| label: fig-evidence-tiers
#| fig-cap: "Evidence tiering of candidate eGenes (placeholder)"

evidence <- tibble(
  tier = c("A: Convergent", "B: Epigenetic-only", "C: Genetic-only"),
  n = c(14, 36, 22)
)

ggplot(evidence, aes(tier, n, fill = tier)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#2b83ba")) +
  labs(x = NULL, y = "Number of genes") +
  theme_minimal()
```

## Session info

```{r session-info}
sessionInfo()
```






















