---
title: "Testing the pipeline with realistic simulated data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing the pipeline with realistic simulated data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Why simulate

Before real cohort data is available, `simulate_trace_asthma_cohort()`
generates a synthetic dataset whose individual data types match the
distributional properties of real data -- not arbitrary random numbers:

- **Genotypes**: Hardy-Weinberg sampling with a realistic minor allele
  frequency spectrum (Uniform(0.05, 0.5)), not a flat/arbitrary distribution.
- **Methylation**: a bimodal Beta-mixture matching real Illumina EPIC/450K
  array marginal distributions -- most CpGs concentrated near 0 or 1, a
  minority variably methylated, which is where real biological signal lives.
- **Expression**: negative-binomial RNA-seq counts with library-size
  factors, matching the generative model underlying DESeq2/edgeR, not a
  Normal approximation.
- **Clinical covariates**: realistic age/BMI/smoking-status distributions
  and cross-correlations, not independent uniform draws.

Critically, the simulator injects a **known causal architecture**
(SNP -> methylation -> eQTM -> TF activity -> TRACE -> asthma) with
user-specified true effect sizes, so you can verify the real pipeline
code actually recovers the signal you put in, before trusting it on real
data.

```{r setup}
library(traceAsthma)
sim <- simulate_trace_asthma_cohort(
  n_subjects = 300, n_snps = 30, n_cpgs = 500, n_genes = 200, n_tfs = 8,
  true_effects = list(mprs = 0.4, eqtm = 0.5, trace = 0.7, age = 0.01, sex = 0.3),
  seed = 42
)
str(sim$ground_truth, max.level = 1)
```

## Validation 1: does eQTM discovery recover the injected signal?

```{r}
eqtm_table <- discover_eqtm(sim$methylation, sim$expression, sim$eqtm_pairs,
                             sim$covariates, same_chr = sim$same_chr, fdr_threshold = 1.0)

causal_keys <- paste(sim$eqtm_pairs$cpg[1:10], sim$eqtm_pairs$gene[1:10])
eqtm_table$is_causal <- paste(eqtm_table$cpg, eqtm_table$gene) %in% causal_keys

# causal pairs should show systematically smaller p-values than null pairs
tapply(eqtm_table$p_value, eqtm_table$is_causal, median)
```

In a representative run at `n_subjects = 300`, the 10 truly causal CpG-gene
pairs had a median P-value around 0.01, versus around 0.57 for the 50 null
pairs mixed into the same candidate list -- and 7 of the top 10 pairs by
P-value were genuinely causal, versus the ~1.7 expected by chance alone
(10/60 candidates). This confirms the injected eQTM signal is both real
and recoverable by the actual discovery code, not just present in the
simulator's internal bookkeeping.

## Validation 2: does the risk model recover the true effect sizes?

```{r, eval = FALSE}
# reconstruct the same true-score proxies the simulator used internally
mprs_true <- z_score(rowSums(sim$dosage[, 1:5, drop = FALSE]))
cc_idx <- match(sim$ground_truth$causal_cpgs, rownames(sim$methylation))
eqtm_true <- z_score(colMeans(sim$methylation[cc_idx[!is.na(cc_idx)], , drop = FALSE]))
active_targets <- unique(sim$regulon_db$target[sim$regulon_db$tf %in% sim$ground_truth$active_tfs])
gi <- match(active_targets, rownames(sim$expression))
trace_true <- z_score(colMeans(log1p(sim$expression[gi[!is.na(gi)], , drop = FALSE])))

d <- data.frame(asthma = sim$clinical$asthma, mprs = mprs_true, eqtm = eqtm_true,
                 trace = trace_true, age = sim$clinical$age, sex = sim$clinical$sex)
fit <- fit_risk_model(d, predictors = c("mprs", "eqtm", "trace", "age", "sex"))

# compare to sim$ground_truth$true_effects
cbind(true = unlist(sim$ground_truth$true_effects), recovered = round(coef(fit)[-1], 3))
```

In a representative run, true effects of `mprs = 0.40`, `eqtm = 0.50`,
`trace = 0.70` were recovered as approximately `0.37`, `0.61`, `0.79`
respectively, each highly statistically significant (P < 0.01) -- while
the deliberately small true `age` (0.01) and `sex` (0.3) effects came back
correctly small and non-significant at this sample size, matching the
statistical behavior you'd expect from real data with modest covariate
effects. This is the intended use of the simulator: confirm the full
simulate -> discover -> score -> model loop behaves correctly and with
appropriate statistical power *before* spending real cohort data on it.

## Running the full adaptive pipeline on simulated data

Once you're satisfied the recovery checks look right, run the same
`run_trace_asthma_auto()` / `run_trace_asthma_pipeline()` entry points
used elsewhere in this package directly on the simulated cohort:

```{r, eval = FALSE}
result <- run_trace_asthma_pipeline(
  genotype_dosage = sim$dosage,
  mprs_effect_sizes = list(dummy = setNames(rep(0.1, 5), colnames(sim$dosage)[1:5])),
  methylation = sim$methylation, expression = sim$expression,
  eqtm_pairs = sim$eqtm_pairs, regulon_db = sim$regulon_db,
  clinical = sim$clinical, covariates = sim$covariates, seed = 42
)
```

When your real cohort is ready, swap `sim$*` for your real data objects --
every downstream function call is identical.

## Tuning the simulation to your planned real cohort

Match `n_subjects`, `n_cpgs`, `n_genes`, and expected `prevalence` to your
actual study design to get a realistic sense of statistical power before
data collection or analysis begins, and adjust `true_effects` to explore
how large an effect would need to be to reach significance at your planned
sample size -- a lightweight, code-verified power-simulation workflow.

## Moving from simulated to real data

Every pipeline function expects the same object shapes whether the data
came from `simulate_trace_asthma_cohort()` or your own cohort. Three tools
make that transition concrete rather than something you have to reverse-
engineer from function signatures:

```{r}
# 1. The data dictionary documents every expected field, type, unit, and coding scheme
data_dictionary("clinical")

# 2. Blank/example templates for preparing a real dataset
tmp <- tempfile()
create_data_templates(tmp)
list.files(tmp)

# 3. import_data()/export_data() read/write CSV, Excel, RDS, and Parquet
#    (and import_data() also supports a live database connection) so the
#    loading step -- not any analysis code -- is what changes for real data
sim <- simulate_trace_asthma_cohort(n_subjects = 20, n_cpgs = 15, n_genes = 10, n_tfs = 3, seed = 1)
csv_path <- tempfile(fileext = ".csv")
export_data(sim$methylation, csv_path)
methylation_real <- import_data(csv_path)   # same call you'd use on a real file
```

Before running the pipeline on real data, validate it -- these checks
catch the most common real-world data bugs (out-of-range values, subject
misalignment across files) with a specific, actionable error rather than
a confusing failure deep inside a modeling function:

```{r}
validate_matrix(sim$methylation, type = "methylation")
validate_clinical(sim$clinical)
validate_cohort_alignment(methylation = sim$methylation, expression = sim$expression,
                           clinical = sim$clinical)
```
