---
title: "The TRACE-Asthma pipeline, end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The TRACE-Asthma pipeline, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette walks through the full TRACE-Asthma pipeline on the tiny,
fully synthetic `example_cohort` dataset bundled with the package (5
patients, 2 SNPs, 2 CpGs, 4 genes, 3 transcription factors). The dataset
exists purely so every stage runs in seconds and every intermediate
number can be hand-checked -- it is **not** sized or powered to produce a
meaningful eQTM discovery or risk model, and none of the numbers below
should be interpreted biologically.

```{r setup}
library(traceAsthma)
data(example_cohort)
str(example_cohort, max.level = 1)
```

## Stage 1: quality control

Genotype, methylation, and RNA-seq QC operate on precomputed summary
statistics (call rates, detection P-values, HWE P-values) rather than
raw IDAT/FASTQ/PLINK files, so that `traceAsthma` has no hard dependency
on the (very large) upstream genomics toolchain. See `?qc_genotype`,
`?qc_methylation`, `?qc_rnaseq` for the expected inputs; a recommended
shell-level upstream pipeline (PLINK2, minfi, STAR/featureCounts) is
documented in `vignette("upstream-pipeline")`.

```{r}
keep_genes <- qc_rnaseq(example_cohort$expression, min_count = 50, min_samples = 2)
keep_genes
```

## Stage 2: eQTM discovery

```{r}
eqtm_table <- discover_eqtm(
  methylation = example_cohort$methylation,
  expression  = example_cohort$expression,
  pairs       = example_cohort$eqtm_pairs,
  covariates  = data.frame(age = example_cohort$covariates$age), # small n -> few covariates
  same_chr    = example_cohort$same_chr,
  fdr_threshold = 1.0  # relaxed threshold; this toy dataset has essentially no power
)
eqtm_table
```

## Stage 3: asthma-relevant filtering and mediation

```{r}
filtered <- filter_asthma_relevant(
  eqtm_table, example_cohort$methylation, example_cohort$expression,
  example_cohort$clinical$asthma,
  data.frame(age = example_cohort$covariates$age),
  p_threshold = 1.0
)
filtered
```

Mediation testing (`test_mediation()`) and elastic-net-based scoring
(`compute_eqtm_score()`, `compute_trace_score()`, `compute_mprs()`)
require the `mediation` and `glmnet` packages respectively; install them
with `install.packages(c("mediation", "glmnet"))` to run the remaining
stages. They are intentionally listed in `Suggests` rather than
`Imports` so that the lightweight QC/discovery/prediction functions above
can be used without pulling in the full modeling stack.

## Stage 4: regulatory network and TF activity

```{r}
regulons <- build_regulons(example_cohort$regulon_db, confidence_classes = c("A", "B", "C"))
enrichment <- test_regulon_enrichment(
  regulons, target_genes = c("IL4", "IL5", "IL13"),
  background_genes = rownames(example_cohort$expression), fdr_threshold = 1.0
)
enrichment
```

## Stages 5-9: TRACE score, MPRS, mediation SEM, integrated model, validation

These stages follow the same pattern and are documented individually:
`?compute_trace_score`, `?compute_mprs`, `?fit_mediation_sem`,
`?fit_risk_model`, `?compare_nested_models`, `?nested_cv`,
`?validate_model`, `?decision_curve`. For a full real-cohort run, the
single entry point is:

```r
result <- run_trace_asthma_pipeline(
  genotype_dosage = NULL,          # omit to skip the MPRS layer
  methylation = methylation,       # CpGs x subjects
  expression = expression,         # genes x subjects
  eqtm_pairs = eqtm_pairs,
  regulon_db = regulon_db,
  clinical = clinical,             # must include an `asthma` (0/1) column
  covariates = covariates,
  seed = 42
)
```

`result$deployable_model` is then ready for [`saveRDS()`] and deployment
via [`predict_asthma_risk()`] -- see `vignette("clinical-deployment-notes")`.

## Predicting risk for a new patient

Once you have a fitted `deployable_model` (either from
`run_trace_asthma_pipeline()` or assembled manually with
`build_deployable_model()`), scoring a new patient requires only their
methylation and expression data:

```r
new_risk <- predict_asthma_risk(
  model = deployable_model,
  methylation = new_patient_methylation,   # CpGs x 1 patient
  expression  = new_patient_expression,    # genes x 1 patient
  clinical    = data.frame(age = 34, sex = "F", bmi = 24.1, smoking = "never")
)
new_risk
```

The returned data frame includes the intermediate `eqtm_score` and
`trace_score` alongside `predicted_probability`, so a clinician or
genetic counselor reviewing the result can see which molecular layer is
driving a given patient's risk estimate rather than only the final
number.

## Adapting to whatever data you actually have

Most real datasets don't include every modality in the full design.
`run_trace_asthma_auto()` inspects what you supply -- at minimum
methylation, expression, and an asthma outcome -- and builds the
best-supported model, skipping the MPRS layer if you have no genotype
data and the TRACE layer if you have no regulon database:

```{r}
# e.g. a user with only CpG data, gene expression, asthma status, age and sex
clinical_partial <- example_cohort$clinical[, c("patient_id", "asthma", "age", "sex")]

plan <- detect_available_data(
  methylation = example_cohort$methylation,
  expression  = example_cohort$expression,
  clinical    = clinical_partial,
  eqtm_pairs  = example_cohort$eqtm_pairs
  # no genotype_dosage, no mprs_effect_sizes, no regulon_db
)
plan
```

`run_trace_asthma_auto()` runs this detection automatically and proceeds
with the maximal model your data supports -- call it exactly like
`run_trace_asthma_pipeline()`, just without needing to specify which
layers to skip.

## Checking what you need to install

```{r}
list_dependencies()
```

Every stage up through eQTM discovery, regulon filtering/enrichment, the
final logistic risk model, and all table/figure export functions run on
base R alone; `glmnet`, `mediation`, `lavaan`, and `viper`/`decoupleR`
are only needed for elastic-net scoring, mediation testing, structural
equation modeling, and TF activity inference respectively, and are
requested with an actionable install message only when actually needed.

## Publication-quality figures and tables

Every major output has a matching Nature-style export function: 600 dpi
figures with direct, non-generic labeling, and fully-labeled, footnoted
tables.

```{r, eval = FALSE}
# Figures (600 dpi, sans-serif, colorblind-safe palette)
plot_nested_auc(result$nested_comparison, path = "figure2_nested_auc.tiff")
plot_trace_weights(result$trace_fit, path = "figure3_trace_weights.png")
plot_mediation_path(sem_result, path = "figure4_mediation.pdf")  # vector, no rasterization

# Tables (fully labeled headers, footnoted abbreviations)
tbl1 <- format_nested_model_table(result$nested_comparison)
export_table(tbl1, "table1_nested_models.csv")

tbl2 <- format_coefficient_table(
  result$risk_model,
  variable_labels = c(eqtm = "Methylation eQTM score (per SD)",
                       trace = "TRACE score (per SD)", age = "Age (years)")
)
export_table(tbl2, "table2_coefficients.csv")
```

## Translating statistics into clinical relevance

Odds ratios and AUC answer "is this associated" and "does this
discriminate." `interpret_clinical_relevance()` answers the question a
clinician actually asks: how much does this change a specific patient's
estimated risk, can that number be trusted, and is there a modifiable
step in the pathway. It computes absolute predicted risk at low/median/
high values of a score (holding other covariates at their sample
median), and layers in a calibration trustworthiness statement and a
mediation-based mechanistic statement when those results are supplied:

```{r, eval = FALSE}
interp <- interpret_clinical_relevance(
  risk_model = result$risk_model,
  data = result$model_data,
  validation = validate_model(result$model_data$asthma,
                               predict(result$risk_model, type = "response")),
  sem_result = sem_result,   # from fit_mediation_sem()
  outcome_label = "asthma"
)
print(interp)
```
